\chapter{Introduction}
\label{sec:introduction}

% TODO: Write introduction

\section{Motivation}

Hierarchical decision-making is a fundamental challenge in many real-world systems. Consider a production facility that must decide which types of jobs to accept within a limited budget, knowing that a scheduler will later assign these jobs to machines to minimize completion time. The facility manager needs to plan for worst-case scenarios: what is the maximum makespan that could occur if jobs were selected adversarially to stress the system, even when the scheduler optimally minimizes completion time? Understanding this worst-case bound is crucial for capacity planning, resource provisioning, and ensuring the facility can meet service level agreements under any job selection within budget.

This scenario exemplifies a \emph{bilevel optimization problem}, where one decision-maker (the leader) makes choices that affect another decision-maker (the follower), who then optimizes their own objective. Such problems could arise in supply chain management, network design, resource allocation, and competitive markets.

\section{Problem Overview}

We study a bilevel optimization problem combining two classic research problems of mathematical optimization:
\begin{itemize}
    \item \textbf{Upper level (Leader):} Knapsack problem -- select job types within a budget
    \item \textbf{Lower level (Follower):} Scheduling problem -- assign jobs to machines to minimize makespan
\end{itemize}

The leader's objective is to maximize the makespan (completion time).
This models a worst-case robust optimization approach, where the leader tries to account for the worst possible makespan,
 even when the follower actually minimizes completion time by finding an optimal schedule. The lower-level scheduling problem is solved via integer programming using Gurobi~\cite{gurobi}.

\section{Contributions}

Throughout this thesis, we refer to this problem as the \emph{bilevel knapsack-scheduling problem}, which concisely describes the combination of knapsack-based job selection at the upper level and parallel machine scheduling at the lower level with adversarial job selection.

This thesis makes the following contributions:
\begin{enumerate}
    \item Formal mathematical model of the bilevel knapsack-scheduling problem
    \item Branch-and-bound algorithm with two different bounds based on a knapsack scheduling relaxation
    \item Efficient dynamic programming solver for the two bounding subproblems
    \item Comprehensive computational study demonstrating practical scalability comparing the two different approaches
    \item Analysis of pruning effectiveness of the bounds and insights into instance characteristics affecting performance
\end{enumerate}

\section{Organization}

The remainder of this thesis is organized as follows. Section~\ref{sec:formulation} presents the mathematical formulation of the bilevel problem. Section~\ref{sec:methodology} describes the branch-and-bound algorithm and bounding techniques. Section~\ref{sec:implementation} discusses implementation details. Section~\ref{sec:experiments} reports computational results. Section~\ref{sec:conclusion} concludes with a summary and future research directions.
