\chapter{Problem Formulation}
\label{sec:formulation}

% TODO: Write problem formulation

\section{Bilevel Optimization Framework}

A bilevel optimization problem has a hierarchical structure with two decision-makers.
\begin{definition}[Bilevel Optimization Problem]
\label{def:bilevel}
\begin{align}
\max_{x} \quad & F(x, y^*(x)) \label{eq:bilevel-upper}\\
\text{s.t.} \quad & x \in X \nonumber\\
\text{where } y^*(x) \in & \arg\min_{y} \{f(x, y) : y \in Y(x)\} \label{eq:bilevel-lower}
\end{align}
\end{definition}

The leader chooses $x$ to optimize their objective $F$, anticipating that the follower will respond by solving their own optimization problem to find $y^*(x)$.  For more information, see~\cite{kleinert2021survey} for a survey on the field of bilevel optimization, and~\cite{dempe2020bilevel} for a comprehensive mathematical foundation.

\textbf{Optimistic vs. Pessimistic Formulation:} When the follower's problem has multiple optimal solutions, 
the \emph{optimistic formulation} assumes the follower selects the solution most favorable to the leader, 
while the \emph{pessimistic formulation} assumes the least favorable choice. The above formulation represents the optimistic case.
\section{The Knapsack-Scheduling Problem}

\subsection{Problem Data}

\begin{itemize}
    \item $n$ job types, indexed by $i = 1, \ldots, n$
    \item Job type $i$ has:
    \begin{itemize}
        \item Processing time (duration): $d_i > 0$
        \item Cost (price): $p_i > 0$
    \end{itemize}
    \item $m$ identical parallel machines
    \item Total budget: $B > 0$
\end{itemize}

\subsection{Leader's Problem (Upper Level)}

The leader selects how many jobs of each type to purchase:

\textbf{Decision variables:} $x_i \in \mathbb{Z}_+$ = number of jobs of type $i$ to select

\textbf{Constraints:}
\begin{equation}
\sum_{i=1}^{n} p_i x_i \leq B \quad \text{(budget constraint)}
\end{equation}

\textbf{Objective:} Maximize the makespan resulting from optimal scheduling:
\begin{equation}
\max_{x} \quad C_{\max}^*(x)
\end{equation}
where $C_{\max}^*(x)$ is the optimal makespan for the follower's problem given selection $x$.

\subsection{Follower's Problem (Lower Level)}

Given the leader's selection $x = (x_1, \ldots, x_n)$, the follower has a total of $\sum_{i=1}^{n} x_i$ jobs to schedule.

\textbf{Decision variables:} For each job $j$, assign it to some machine

\textbf{Objective:} Minimize makespan (maximum machine load)
\begin{equation}
C_{\max}^*(x) = \min \max_{k=1,\ldots,m} L_k
\end{equation}
where $L_k$ denotes the load on machine $k$, defined as the total processing time of all jobs assigned to machine $k$ after scheduling.

\section{Complete Bilevel Formulation}

We present a mixed-integer linear programming (MILP) formulation compatible with standard bilevel solvers such as MibS~\cite{Tahernejad2020}. The formulation uses a shared makespan variable $C_{\max}$ that appears in both levels:

\begin{align}
\max_{x} \quad & C_{\max} \label{eq:leader-obj}\\
\text{s.t.} \quad & \sum_{i=1}^{n} p_i x_i \leq B \label{eq:budget}\\
& x_i \in \mathbb{Z}_+ \quad \forall i \in \{1, \ldots, n\} \nonumber\\[1ex]
\text{where } C_{\max} = \quad & \min_{y, \hat{C}_{\max}} \quad \hat{C}_{\max} \label{eq:follower-obj}\\
\text{s.t.} \quad & \sum_{k=1}^{m} y_{ik} = x_i \quad \forall i \in \{1, \ldots, n\} \label{eq:assign}\\
& \hat{C}_{\max} \geq \sum_{i=1}^{n} d_i y_{ik} \quad \forall k \in \{1, \ldots, m\} \label{eq:makespan}\\
& y_{ik} \in \mathbb{Z}_+ \quad \forall i, k \nonumber
\end{align}

The leader's variables are $x_i$ (number of jobs of type $i$ to select) and the follower's variables are $y_{ik}$ (number of jobs of type $i$ assigned to machine $k$). The variables $C_{\max}$ and $\hat{C}_{\max}$ represent the makespan: the leader seeks to maximize it while the follower seeks to minimize it, creating the adversarial nature of the problem.

\section{Problem Characteristics}

The computational challenge of this problem arises from multiple sources. First, the problem combines two NP-hard problems (knapsack and scheduling) in a bilevel scenario. Bilevel optimization is well-known to be computationally challenging even for simple problems with linear objectives and constraints, making our discrete, non-convex variant particularly difficult to solve. The follower's optimal value function $C_{\max}^*(x)$ is non-convex and discontinuous, which prevents the use of gradient-based methods and complicates the analysis of the leader's objective landscape. Furthermore, the discrete nature of both levels means that standard continuous optimization techniques, such as KKT conditions, cannot be applied. Consequently, solving this problem requires combinatorial methods that explicitly explore the discrete solution space.

\section{Interpretation: Why Maximize Makespan?}

The leader maximizes makespan to identify the worst-case completion time that could arise within the budget constraint. This defensive approach models robust planning under uncertainty, where the decision-maker seeks to understand the maximum possible makespan that could result from any job selection, even when the scheduler optimizes to minimize completion time. By solving for this worst-case bound, the leader can make informed decisions about resource provisioning, capacity planning, and service level guarantees. More generally, this framework applies to any adversarial or worst-case planning scenario where a decision-maker must prepare for unfavorable conditions while an opposing agent optimizes in their own interest.

\textbf{Example:} A production facility manager needs to determine what workload the system must be prepared to handle. By identifying the job selection that maximizes makespan (under optimal scheduling), the manager obtains a worst-case guarantee: the system will never face a longer completion time than this bound, regardless of which specific jobs arrive within budget.

\section{Literature Review}

We provide a brief overview of selected works relevant to our problem; for a comprehensive treatment of bilevel optimization, we refer the reader to the extensive body of literature in this field. 

Dempe and Zemkoho~\cite{dempe2020bilevel} provide a comprehensive mathematical foundation for bilevel optimization, covering theory, optimality conditions, and solution approaches. Kleinert et al.~\cite{kleinert2021survey} offer an extensive survey on mixed-integer programming techniques in bilevel optimization, reviewing branch-and-bound methods, cutting planes, and decomposition approaches that form the algorithmic foundation for modern bilevel solvers.


Most relevant to our work, Weninger and Fukasawa~\cite{weninger2025bilevel} recently developed a fast combinatorial branch-and-bound algorithm for the bilevel knapsack problem with interdiction constraints. 
Their approach uses dynamic programming to compute strong lower bounds by solving a relaxed 2n-level version of the problem where items are processed in an online fashion. 
This technique shares similarities with our method, as both employ branch-and-bound combined with dynamic programming to exploit problem structure. 
However, their lower-level problem is a second knapsack optimization problem, whereas ours involves parallel machine scheduling, requiring different bounding techniques.

Leitner et al.~\cite{leitner2023exact} further advanced the field by developing exact methods for three-level optimization problems (fortification games), demonstrating that problem-specific reformulations and specialized cuts can significantly outperform general approaches, solving instances with over 330,000 nodes to optimality.

On the algorithmic side, Tahernejad, Ralphs, and DeNegre~\cite{Tahernejad2020} developed MibS, 
a general-purpose solver for mixed-integer bilevel linear optimization problems. 
MibS implements a comprehensive branch-and-cut algorithm that combines features from traditional MILP and bilevel methods with new techniques to create a flexible and robust solver.
Fischetti et al.~\cite{fischetti2016intersection} used MibS as a benchmark to compare their proposed intersection cuts for Mixed-Integer Bilevel Linear Programs, 
demonstrating that problem-specific cutting planes can significantly improve performance over general-purpose solvers. 
In this thesis, we follow their approach by comparing our specialized branch-and-bound algorithm against MibS that is open source and available for public use to evaluate the benefits of problem-specific techniques versus general-purpose bilevel solvers.

While the general bilevel optimization literature is extensive, our specific problem combining knapsack constraints in the upper level with parallel machine scheduling in the lower level represents a novel variant. The hierarchical structure with adversarial objectives (maximize vs. minimize makespan) distinguishes our problem from classical bilevel scheduling formulations, motivating the development of tailored solution methods that exploit the specific structure of the knapsack-scheduling problem.
