\chapter{Conclusion and Future Work}
\label{sec:conclusion}

\section{Summary and Main Contributions}

This thesis investigated the Bilevel Parallel Machine Scheduling problem with Adversarial Job Selection, where a leader selects job types within a budget to maximize makespan, anticipating that a follower will optimally schedule jobs to minimize completion time. The problem combines two classic optimization challenges—knapsack selection and parallel machine scheduling—within a hierarchical, adversarial framework.

The main contributions are:
\begin{itemize}
    \item Formal mathematical model for the bilevel knapsack-scheduling problem, including leader and follower objectives and constraints.
    \item Development of a branch-and-bound algorithm with efficient pruning, utilizing both ceiling knapsack and MaxLPT bounds, and incorporating an early optimality detection mechanism that can solve certain instances directly at the root node.
    \item Implementation of dynamic programming for bound computation and incremental LPT scheduling for fast makespan evaluation.
    \item Integration of Gurobi~\cite{gurobi} for solving integer programming subproblems, ensuring optimality in the follower's schedule.
    \item Comprehensive computational study demonstrating practical scalability, pruning effectiveness, and correctness across a wide range of instance types and parameters, including comparisons to the general-purpose MibS solver.
\end{itemize}

\section{Key Results and Insights}
We compared our algorithm against MibS~\cite{Tahernejad2020}, and while our method significantly outperforms MibS on our specialized problem, 
it is important to note that MibS is a general-purpose solver designed for a far broader class of bilevel optimization problems.
The tailored branch-and-bound algorithm for knapsack-scheduling problems achieves substantial speedups over general-purpose solvers, successfully solving instances with up to 26 job types and 12 machines, 
and demonstrating scalability beyond these sizes in certain cases. In the following, we summarize the key points from the algorithm's design and performance:
\begin{itemize}
    \item The LPT rule provides a $\frac{4}{3}$-approximation for makespan bounds, guaranteeing good starting incumbents.
    \item Tight bounds from the ceiling bound and LPT bound enable effective pruning, reducing the search space and improving efficiency. 
    Compared to the MibS solver, the median of the speed difference is almost the entire time limit of 3600s, 
    showing that our method can solve instances in fractions of a second that MibS cannot solve within the time limit.
    \item Compared to MibS, the maxLPT bound was able to solve 179 out of 180, while MibS was only able to solve 35 out of 180 instances within the time limit, demonstrating the significant advantage of problem-specific bounding techniques.
    \item The early optimality detection mechanism allows the algorithm to solve certain instances directly at the root node, significantly reducing runtime for those cases.
    \item The modular implementation supports efficient node expansion, incremental updates, and memory savings compared to breadth-first approaches.
    \item Gurobis~\cite{gurobi} integration ensures that the follower's scheduling problem is solved optimally using state-of-the-art integer programming techniques, contributing to the overall performance of the algorithm.
    \item Computational experiments confirm the algorithm's robustness across diverse instance types, including uniform, high variance, and correlated job structures.
\end{itemize}

\section{Limitations}
With our advances in modeling and solving bilevel knapsack-scheduling problems, there are still several limitations to address in future work:
\begin{itemize}
    \item There is currently no theory guaranteeing for which instances the optimal solution can be found at the root node.
    \item Our branch-and-bound algorithm is designed for the specific structure of the knapsack-scheduling problem and not directly applicable to other bilevel problems without significant modifications.
    \item Scalability is limited for very large instances (e.g., 30+ job types or extreme budget/machine counts) due to exponential complexity if the solution cannot be obtained at the root node.
    \item The current model focuses solely on makespan; real-world applications may require multi-objective optimization (cost, energy, risk distribution).
    \item The algorithm is designed for parallel machine scheduling; extensions to unrelated machines or other scheduling environments may require significant modifications.
\end{itemize}

\section{Future Directions}

The advances achieved in modeling and solving bilevel knapsack-scheduling problems in this thesis highlight several opportunities for continued research and development:
\begin{itemize}
    \item Algorithmic improvements: Enhanced pruning, symmetry breaking and parallelization.
    \item Theoretical analysis: Characterization of instance properties that allow optimal solutions at the root node.
    \item Problem extensions: further constraints, stochastic processing times, and multi-objective formulations.
    \item Unrelated machines could be an interesting extension, as ceiling bound may be easily applicable to this setting. While the LPT bound outperforms the ceiling bound in the parallel machine setting, the relatively straightforward approach of the ceiling bound makes it a promising candidate for the unrelated machine setting.
    \item Application areas: Production scheduling, capacity planning, project management, and resource allocation under uncertainty.
\end{itemize}


\section{Closing Remarks}

This thesis addressed the Bilevel Parallel Machine Scheduling problem with Adversarial Job Selection, 
a challenging hierarchical optimization problem that integrates leader-level unbounded knapsack selection with follower-level optimal scheduling. 
The work provided a formal mathematical model, clarified the adversarial structure, and analyzed the inherent computational complexity of the problem.

The main contribution is the development of a dedicated branch-and-bound algorithm that leverages both ceiling knapsack and MaxLPT bounds for effective pruning. 
The algorithm incorporates an early optimality detection mechanism, dynamic programming for bound computation, 
and incremental LPT scheduling for efficient makespan evaluation. Integration with integer programming solvers ensures that the follower's scheduling problem is solved to optimality at each node.

Comprehensive computational experiments demonstrated that the proposed approach achieves substantial speedups over state-of-the-art methods like MibS,
 scales to problem sizes beyond the reach of naive methods, and robustly handles a wide range of instance types and parameter settings. 
 The results highlight the effectiveness of tight bounding and principled search in reducing the search space and enabling the solution of complex adversarial scheduling problems that were previously intractable.

Overall, this work establishes a rigorous algorithmic and computational foundation for bilevel knapsack-scheduling problems, providing new insights into their structure and tractability. 
